---
title: "Test Weibull Z"
author: "Joanne"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Weibull Test : in theory

## Test 

$$
\mathcal{H}_0 : \lambda(.) \text{ constant} \quad \quad \mbox{contre} \quad \quad \mathcal{H}_1^+ : \lambda(.) \text{ increasing} \\
\mathcal{H}_0 : \beta=1 \quad \quad \mbox{contre} \quad \quad \mathcal{H}_1^+ : \beta > 1
$$

with $\lambda(t) = \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}$

## The test statistic: Z

$$
Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right )
$$

Reject $\mathcal{H}_0$ when : $\mathcal{R} = \{Z \leq q \}$.

When Z take smaller values, it means that the ratios $(T^\star/T_i)_i$ are smaller. So the $(T_i)_i$ are higher and clothe to $T^\star$, which appends when $\lambda(.)$ is increasing. When $\lambda(.)$ is constant, the gap between the $(T_i)_i$ is stable, but when $\lambda(.)$ is increasing, the gap between the $(T_i)_i$ gets smaller and smaller.

## The law

According to the paper, we have : 

$$
\Lambda(T_k) / \Lambda(T^\star) = U_{(k)}
$$
where $(U_{(i)})_i$ order statistics generated from a random sample from $\mathcal{U}(0,1)$.

$$
\frac{\int_0^{T_k} \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}dt}{\int_0^{T^\star} \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}dt} = U_{(k)} \\
\frac{(T_k/\theta)^\beta}{(T^\star/\theta)^\beta} = \left(\frac{T_k}{T^\star} \right)^\beta = U_{(k)}
$$

So $\frac{T_k}{T^\star}$ has the same distribution as $(U_{(k)})^{1/\beta}$.

## 1 - Simulation

Loi des $W_i$ : exponentielle 

```{r,eval=FALSE}
simulPPh2 <- function(lambda,n)
{
  W <- rexp(n, rate = lambda) 
  Ti <- cumsum(W)
  return(Ti)
}
```

```{r,eval=FALSE}
n <- 20
lambda <- 2
PPh2 = simulPPh2(lambda = 2, n = 20)

# plot the counting process (with jumps = 1): 
plot(c(0,PPh2),seq(0,length(PPh2)),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times: 
points(PPh2, rep(0,length(PPh2)),type="p")

# link the arrival times with the counts:
lines(PPh2, seq(1,length(PPh2)),type="h",lty=2)
```

# 3 - Inhomogeneous Poisson processes

```{r,eval=FALSE}
simulPPi = function(lambda_fct,Tmax,M)
{
  Ti <- simulPPh1(M, Tmax)
  n <- length(Ti)
  U <- runif(n, min = 0, max = M)
  ti <- c()
  for (i in 1:n){
    if (U[i] < lambda_fct(Ti[i])) {
      ti <- c(ti, Ti[i])
    }
  }
  return(sort(ti))
}
```

```{r,eval=FALSE}
Tmax=10
lambda_fct1 <- function(t){
  return(ifelse(t <= 7, 2, 
                ifelse(t >8, 8, 0)))
  }
M1=10

PPi1 = simulPPi(lambda_fct1, Tmax, M1)

# plot the counting process (with jumps = 1): 
plot(c(0,PPi1),seq(0,length(PPi1)),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times: 
points(PPi1, rep(0,length(PPi1)),type="p")

# link the arrival times with the counts:
lines(PPi1, seq(1,length(PPi1)),type="h",lty=2)
```

```{r,eval=FALSE}
Tmax=10
lambda_fct2 <- function(t){
  return(2*t)
  }
M2= 21

PPi2 = simulPPi(lambda_fct2, Tmax, M2)

# plot the counting process (with jumps = 1): 
plot(c(0,PPi2),seq(0,length(PPi2)),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times: 
points(PPi2, rep(0,length(PPi2)),type="p")

# link the arrival times with the counts:
lines(PPi2, seq(1,length(PPi2)),type="h",lty=2)
```




