---
title: "Test Weibull Z"
author: "Joanne"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Weibull Test : in theory

## Test 

$$
\mathcal{H}_0 : \lambda(.) \text{ constant} \quad \quad \mbox{contre} \quad \quad \mathcal{H}_1^+ : \lambda(.) \text{ increasing} \\
\mathcal{H}_0 : \beta=1 \quad \quad \mbox{contre} \quad \quad \mathcal{H}_1^+ : \beta > 1
$$

with $\lambda(t) = \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}$

## The test statistic: Z

$$
Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right )
$$

Reject $\mathcal{H}_0$ when : $\mathcal{R} = \{Z \leq q \}$.

When Z take smaller values, it means that the ratios $(T^\star/T_i)_i$ are smaller. So the $(T_i)_i$ are higher and clothe to $T^\star$, which appends when $\lambda(.)$ is increasing. When $\lambda(.)$ is constant, the gap between the $(T_i)_i$ is stable, but when $\lambda(.)$ is increasing, the gap between the $(T_i)_i$ gets smaller and smaller.

## The law

According to the paper, we have : 

$$
\Lambda(T_k) / \Lambda(T^\star) = U_{(k)}
$$
where $(U_{(i)})_i$ order statistics generated from a random sample from $\mathcal{U}(0,1)$.

$$
\frac{\int_0^{T_k} \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}dt}{\int_0^{T^\star} \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}dt} = U_{(k)} \\
\frac{(T_k/\theta)^\beta}{(T^\star/\theta)^\beta} = \left(\frac{T_k}{T^\star} \right)^\beta = U_{(k)}
$$

So $\frac{T_k}{T^\star}$ has the same distribution as $(U_{(k)})^{1/\beta}$.

Under $\mathcal{H}_0$, $\frac{T_k}{T^\star}$

# 3 - Inhomogeneous Poisson processes

We simulate $N \sim \mathcal{P}(\Lambda(t))$.

We simulate $S_1, ..., S_N$ i.i.d from density $s \mapsto \frac{\lambda(s)}{\Lambda(T^\star)} \mathbb{1}_{0 < s \leq T^\star}$. To do so, we apply the inverse distribution function or quantile function ($F^{-1}$) to a uniform sample. 

Then $(T_1, ..., T_N) = (S_{(1)}, ..., S_{(N)})$.

We have $\Lambda = (T^\star / \theta)^\beta$.

```{r}
lambda_weibull <- function(t, beta, theta){
  return((beta/theta)*(t/theta)**(beta-1))
}

lambda_int <- function(Tstar, beta, theta){
  return((Tstar / theta)**beta)
}

density_S <- function(s, lambda_fct, lambda_int, Tmax, beta, theta){
  if (s <= Tmax) {
    p <- lambda_fct(s, beta, theta)/lambda_int(Tmax, beta, theta)
  } else {
    p <- 0
  }
  return(p)
}

quantile_function <- function(p, Tstar, beta){
  return(Tstar * p**(1/beta))
}


simulHPP <- function(Tstar, beta, theta){
  # simulate the number of event under a Poisson distribution
  N <- rpois(1, lambda_int(Tstar, beta, theta))
  # simulate a uniform sample
  U <- runif(N,0,1)
  # apply the quantile function (inverse of the cumulative distribution function)
  S <- quantile_function(U, Tstar, beta)
  # sort the sample
  return(sort(S)[1:N])
}
```

Let's take $\theta = 1$.

```{r}
Tstar <- 1

plot(seq(0,Tstar,length.out = 10), lambda_weibull(seq(0,Tstar,length.out = 10), 5, 1), 
     main = "beta > 1 : lambda increasing")

plot(seq(0,Tstar,length.out = 10), lambda_weibull(seq(0,Tstar,length.out = 10), 1, 1), 
     main = "beta = 1 : lambda constant")
```


```{r}
theta <- 0.4
Tstar <- 1


beta <- 2 # (over 1 = increasing)
PPi1 <- simulHPP(Tstar, beta, theta)

plot(c(0,PPi1),seq(0,length(PPi1)),type="s",xlab="time t",ylab="number of events by time t")
points(PPi1, rep(0,length(PPi1)),type="p")
lines(PPi1, seq(1,length(PPi1)),type="h",lty=2)


beta <- 1 # (constant)
N <- 0
while (N == 0) {
  PPi2 <- simulHPP(Tstar, beta, theta)
  N <- length(PPi2)
}

plot(c(0,PPi2),seq(0,length(PPi2)),type="s",xlab="time t",ylab="number of events by time t")
points(PPi2, rep(0,length(PPi2)),type="p")
lines(PPi2, seq(1,length(PPi2)),type="h",lty=2)
```


# Test 

$\beta Z$ follows a chi-squared with $2n$ degrees of freedom.

$$
\mathcal{R} = \{Z \leq q \} \\
\text{with  } Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right ) \\
\text{and } \beta Z \sim \mathcal{X}^2(2n)
$$

P-value computation :

$$
pvalue = \mathbb{P}(Z \leq Z^{obs}) = \Phi_{\mathcal{X^2}}(Z^{obs}) \\
\text{where } \Phi \text{ cdf of a } \mathcal{X}^2(2n)
$$

Let's compute $Z^{obs}$ and the p-value :

```{r}
get_pval <- function(PPi, Tstar){
  Zobs = 2*sum(log(Tstar/PPi))
  N = length(PPi)
  # pvalue : smaller value such that we reject H0
  pval = pchisq(Zobs, 2*N)
  return(pval)
}

alpha <- 0.05

# With PPi1
pval <- get_pval(PPi1, Tstar)
cat("P-value : ", pval, "\n")

# accept
if (pval < alpha) {
  cat("Reject H0 : lambda is inscreasing\n")
} else {
  cat("Accept H0 : lambda is constant\n")
}

# With PPi2
pval <- get_pval(PPi2, Tstar)
cat("P-value : ", pval, "\n")

# accept
if (pval < alpha) {
  cat("Reject H0 : lambda is inscreasing\n")
} else {
  cat("Accept H0 : lambda is constant\n")
}

```

## Size $\alpha$

To check if the test's size is indead $\alpha = 0.05$, we will run it $x$ times, and count the number of time it gets the right answer.


```{r}
get_size <- function(K, alpha, get_pval, simulHPP, Tstar, beta, theta){
  nb.rejects <- 0
  for(k in 1:K){
    N <- 0
    while(N <= 1){
      PPi <- simulHPP(Tstar, beta, theta)
      N <- length(PPi)
    }
    pval = get_pval(PPi, Tstar)
    nb.rejects = nb.rejects + (pval<=alpha)  
  }
  return(nb.rejects/K)
}
```


```{r}
# question : on compte le nombre de fois qu'il dit H1 alors que c'est H0
# mais ducoup on ne le test que avec des lambda constant ?

alpha <- 0.05
beta <- 1
theta <- 1
Tstar <- 2
test_size <- get_size(K = 1000, alpha, get_pval, simulHPP, Tstar, beta, theta)
cat("\nSize of the test : ", test_size)

```

## How to compute the power of the test

The power of the test is $\mathbb{P}_{\mathcal{H}_1}(\text{reject } \mathcal{H}_0)$. So we need to know the law of $Z$ under $\mathcal{H}_1$. Here we have $\beta Z \sim \mathcal{X}^2(2n)$. And we have $\mathcal{R} = \{Z \leq q_\alpha \}$, where $q_\alpha$ the $\alpha$ quantile of $\mathcal{X}^2(2n)$.

$$

$$



```{r}
Tstar <- 2
theta <- 1
beta <- 4
N <- 0
while(N <= 1){
  PPi <- simulHPP(Tstar, beta, theta)
  N <- length(PPi)
}

cat(N, "\n")
pchisq(beta*qchisq(alpha, 2*N), 2*N)
```


```{r}
alpha <- 0.05
beta <- 4
theta <- 1
Tstar <- 2
test_power <- get_size(K = 1000, alpha, get_pval, simulHPP, Tstar, beta, theta)
cat("\nPuissance of the test : ", test_power)


```



