---
title: "Testing for homogeneity of a Poisson process"
author: "Joanne Couallier, Alexandre Personnic, Huimin ZHANG"
date: "07/12/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Presentation of the different test

The aim of this project is to test 
$$
\mathcal{H}_0 : \lambda(.) \text{ constant} \quad \quad \mbox{against} \quad \quad \mathcal{H}_1^+ : \lambda(.) \text{ increasing}
$$

using different test and different function for $\lambda$.

## Laplace test

Test statistic : 

$$
L = \sum_{i=1}^n \frac{T_i}{T^*}
$$
where $T^*$ is a fixed units of time

Reject $\mathcal{H}_0$ when : $\mathcal{R}_{\alpha} = \left\{ L \ge l \right\}$

## Weibull test

Test statistic :

$$
Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right )
$$

Reject $\mathcal{H}_0$ when : $\mathcal{R}_{\alpha} = \{Z \leq q \}$

## F-test

Test statistic : 

$$F=\frac{(n-d)T_d}{d(T_n-T_d)}$$


# Numerical simulations

## Simulation with homogeneous Poisson processses

```{r}
plot_PP <-function(PP){
  # plot the counting process (with jumps = 1): 
  plot(c(0,PP),seq(0,length(PP)),type="s",xlab="time t",ylab="number of events by time t")
  # add the arrival times: 
  points(PP, rep(0,length(PP)),type="p")
  # link the arrival times with the counts:
  lines(PP, seq(1,length(PP)),type="h",lty=2)
}
```

```{r}
simulPPh <- function(lambda, Tstar)
{
  Y <- rpois(1, lambda*Tstar)
  U <- runif(Y, min = 0, max = Tstar)
  return(sort(U))
}
```

### Laplace test

```{r}
# In the homogeneous case, beta = 0, so lambda = alpha
Tstar <- 5 
alpha <- 2
lambda <- alpha
PPh1 <- simulPPh(lambda, Tstar)
plot_PP(PPh1)
```
Test : TO EXPLAIN

```{r}
get_pvalL <- function(PP, Tstar)
{
  n <- length(PP)
  L_obs <- sum(PP/Tstar)
  Z_obs <- (L_obs - n/2)/sqrt(n/12)
  # pvalue : smaller value such that we reject H0
  pval <- 1 - pnorm(Z_obs)
  return(pval)
}

decision <- function(pval, alpha) {
  if (pval < alpha) {
    cat("Reject H0 : lambda is inscreasing\n")
  } else {
    cat("Accept H0 : lambda is constant\n")
  }
}
```

```{r}
alpha <- 0.05
pval <- get_pvalL(PPh1, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```


```{r}
get_size <- function(K, alpha, get_pval, simulPP, lamdba, Tstar){
  nb.rejects <- 0
  for(k in 1:K){
    N <- 0
    while(N <= 1){
      PP <- simulPP(lambda, Tstar)
      N <- length(PP)
    }
    pval <- get_pval(PP, Tstar)
    nb.rejects = nb.rejects + (pval<=alpha)  
  }
  return(nb.rejects/K)
}
```

```{r}
alpha <- 0.05
Tstar <- 5
lambda <- 2
test_size <- get_size(K = 2000, alpha, get_pvalL, simulPPh, lambda, Tstar)
cat("\nSize of the test : ", test_size)
```

### Weibull test

```{r}
# In the homogeneous case, beta = 1, so lambda = 1/theta
Tstar <- 1
theta <- 0.1
lambda <- 1/theta
PPh2 <- simulPPh(lambda, Tstar)
plot_PP(PPh2)
```

$\beta Z$ follows a chi-squared with $2n$ degrees of freedom.

$$
\mathcal{R} = \{Z \leq q \} \\
\text{with  } Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right ) \\
\text{and } \beta Z \sim \mathcal{X}^2(2n)
$$

P-value computation :

$$
pvalue = \mathbb{P}(Z \leq Z^{obs}) = \Phi_{\mathcal{X^2}}(Z^{obs}) \\
\text{where } \Phi \text{ cdf of a } \mathcal{X}^2(2n)
$$

```{r}
get_pvalZ <- function(PP, Tstar){
  Zobs <- 2*sum(log(Tstar/PP))
  N <- length(PP)
  # pvalue : smaller value such that we reject H0
  pval <- pchisq(Zobs, 2*N)
  return(pval)
}
```

```{r}
alpha <- 0.05
pval <- get_pvalZ(PPh2, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

```{r}
alpha <- 0.05
Tstar <- 5
theta <- 0.1
lambda <- 1/theta
test_size <- get_size(K = 2000, alpha, get_pvalZ, simulPPh, lambda, Tstar)
cat("\nSize of the test : ", test_size)
```



### F-test

```{r}
lambda <- 2
Tstar <- 20
PPh3 <- simulPPh(lambda, Tstar)
plot_PP(PPh3)
```

```{r}
get_pvalF <- function(PP, Tstar = 1){ # The parameter Tstar here is useless, we add it so that we can use the generic function get_size defined before 
  n = length(PP)
  d = floor(n/2)
  F = (n-d)*PP[d]/(d*(PP[n] - PP[d]))
  pval = 1. - pf(F, 2*d, 2*(n-d))
  return(pval)
}
```

```{r}
alpha <- 0.05
pval <- get_pvalF(PPh3)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

```{r}
alpha <- 0.05
Tstar <- 20
lambda <- 2
test_size <- get_size(K = 2000, alpha, get_pvalF, simulPPh, lambda, Tstar)
cat("\nSize of the test : ", test_size)
```


## Simulation with inhomogeneous Poisson processses

```{r}
simulPPi <- function(LAMBDA, quantile_function, Tstar){
  # simulate the number of event under a Poisson distribution
  N <- rpois(1, LAMBDA(Tstar))
  # simulate a uniform sample
  U <- runif(N,0,1)
  # apply the quantile function (inverse of the cumulative distribution function)
  S <- quantile_function(U)
  # sort the sample
  return(sort(S)[1:N])
}
```

### Laplace test

```{r}
lambda_exp <- function(alpha, beta, Tstar) {
  lambda <- function(t) {
    return(alpha*exp(beta*t))
  }
  LAMBDA <- function(t) {
    return((alpha/beta)*(exp(beta*t) - 1))
  }
  
  quantile_function <- function(p) {
    return((1/beta) * log(p*(exp(beta*Tstar) - 1) + 1))
  }
  
  return(c(lambda, LAMBDA, quantile_function))
}
```

```{r}
Tstar <- 3
alpha <- 1 
beta <- 1
expo <- lambda_exp(alpha, beta, Tstar)
lambda <- expo[[1]]
LAMBDA <- expo[[2]]
quantile_function <- expo[[3]]
PPi1 <- simulPPi(LAMBDA, quantile_function, Tstar)
plot_PP(PPi1)
```
 
```{r}
alpha <- 0.05
pval <- get_pvalL(PPi1, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```
 Power of the test : 
 
```{r}
get_power <- function(K, alpha, get_pval, simulPP, LAMBDA, quantile_function, Tstar){
  nb.rejects <- 0
  for(k in 1:K){
    N <- 0
    while(N <= 1){
      PP <- simulPP(LAMBDA, quantile_function, Tstar)
      N <- length(PP)
    }
    pval <- get_pval(PP, Tstar)
    nb.rejects = nb.rejects + (pval<=alpha)  
  }
  return(nb.rejects/K)
}
```
 
```{r}
alpha <- 0.05
Tstar <- 3
test_power <- get_power(K = 2000, alpha, get_pvalL, simulPPi, LAMBDA, quantile_function, Tstar)
cat("\nPower of the test : ", test_power)
``` 
 

### Weibull test 

```{r}
lambda_weibull <- function(beta, theta, Tstar) {
  lambda <- function(t) {
    return((beta/theta)*(t/theta)**(beta-1))
  }
  LAMBDA <- function(t) {
    return((t / theta)**beta)
  }
  quantile_function <- function(p) {
    return(Tstar * p**(1/beta))
  }
  return(c(lambda, LAMBDA, quantile_function))
}
```

```{r}
Tstar <- 1
beta <- 3
theta <- 0.5
weib <- lambda_weibull(beta, theta, Tstar)
lambda <- weib[[1]]
LAMBDA <- weib[[2]]
quantile_function <- weib[[3]]
PPi2 <- simulPPi(LAMBDA, quantile_function, Tstar)
plot_PP(PPi2)
```

```{r}
alpha <- 0.05
pval <- get_pvalZ(PPi2, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

```{r}
alpha <- 0.05
Tstar <- 3
beta <- 4
theta <- 1
weib <- lambda_weibull(beta, theta, Tstar)
lambda <- weib[[1]]
LAMBDA <- weib[[2]]
quantile_function <- weib[[3]]
test_power <- get_power(K = 2000, alpha, get_pvalZ, simulPPi, LAMBDA, quantile_function, Tstar)
cat("\nPower of the test : ", test_power)
``` 


### F-test

```{r, eval=TRUE}
# lambda increasing by level, return lambda, LAMBDA, quantile_function
lambda_palier <- function(Tstar=10, tau=0.5, a=1, b=2) {
  # a function returning a if t <= tau and b otherwise.
  lambda <- function(t) {
    return(ifelse(t <= tau, a, b))
  }
  # return a function returning the primitive of the precedent one.
  LAMBDA <- function(t) {
    return(ifelse(t <= tau, a*t, a*tau+b*(t-tau)))
  }
  # return a function returning the inverse of the precedent one.
  quantile_function <- function(k) {
    return(ifelse(k<=tau*a, LAMBDA(Tstar)*k/a, (LAMBDA(Tstar)*k-a*tau)/b + tau))
  }
  return(c(lambda, LAMBDA, quantile_function))
}
```

```{r,eval=TRUE}
# return lambda affine function of t with origin and coefficient
lambda_affine <- function(origin=0, coef=1) {
  if(coef == 0) {
    stop("coef must not be 0.")
  }
  temp_fun <- function(t) {
    return(origin+t*coef)
  }
  return(temp_fun)
}
```

```{r}
Tstar <- 20
tau <- 2/3
a <- 1
b <- 2
palier <- lambda_palier(Tstar, tau*Tstar, a, b)
lambda <- palier[[1]]
LAMBDA <- palier[[2]]
quantile_function <- palier[[3]]
# PPi3 <- simulPPi(palier[[1]], Tmax, palier[[3]])
PPi3 <- simulPPi(LAMBDA, quantile_function, Tstar)
plot_PP(PPi3)
```

```{r}
alpha <- 0.05
pval <- get_pvalF(PPi3, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

```{r}
alpha <- 0.05
test_size <- get_power(K = 2000, alpha, get_pvalF, simulPPi, LAMBDA, quantile_function, Tstar)
cat("\nPower of the test : ", test_size)
``` 

