---
title: "Testing for homogeneity of a Poisson process"
author: "Joanne Couallier, Alexandre Personnic, Huimin ZHANG"
date: "07/12/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The aim of this project is to test 
$$
\mathcal{H}_0 : \lambda(.) \text{ constant} \quad \quad \mbox{against} \quad \quad \mathcal{H}_1 : \lambda(.) \text{ increasing}
$$
using different test and different function for $\lambda$.


# Implementation of general function useful for all the test

```{r}
# Simulation of a homogeneous Poisson process
simulPPh <- function(lambda, Tstar)
{
  Y <- rpois(1, lambda*Tstar)
  U <- runif(Y, min = 0, max = Tstar)
  return(sort(U))
}
```

```{r}
# Simulation of an inhomogeneous Poisson process
simulPPi <- function(LAMBDA, quantile_function, Tstar){
  # simulate the number of event under a Poisson distribution
  N <- rpois(1, LAMBDA(Tstar))
  # simulate a uniform sample
  U <- runif(N,0,1)
  # apply the quantile function (inverse of the cumulative distribution function)
  S <- quantile_function(U)
  # sort the sample
  return(sort(S)[1:N])
}
```

```{r,eval=TRUE}
simulPPi_thining <- function(lambda, placeholder, Tstar) {
  # Process to simulate an inhomogeneous Poisson process
  # with a given lambda function and Tstar using thinning process.
  M = 1.2 * lambda(Tstar)
  Ti <- simulPPh(M, Tstar)
  n <- length(Ti)
  U <- runif(n, min = 0, max = M)
  ti <- c()
  for (i in 1:n){
    if (U[i] < lambda(Ti[i])) {
      ti <- c(ti, Ti[i])
    }
  }
  return(sort(ti))
}
```

```{r}
# Plot a Poisson process
plot_PP <-function(PP){
  # plot the counting process (with jumps = 1): 
  plot(c(0,PP),seq(0,length(PP)),type="s",xlab="time t",ylab="number of events by time t")
  # add the arrival times: 
  points(PP, rep(0,length(PP)),type="p")
  # link the arrival times with the counts:
  lines(PP, seq(1,length(PP)),type="h",lty=2)
}
```

```{r}
# Decide if we reject H0 or not 
decision <- function(pval, alpha) {
  if (pval < alpha) {
    cat("Reject H0 : lambda is inscreasing\n")
  } else {
    cat("Accept H0 : lambda is constant\n")
  }
}
```


# Laplace test 

The test statistic is defined by : 

$$
L = \sum_{i=1}^n \frac{T_i}{T^*}
$$
where $T^*$ is a fixed units of time

This test can be used for testing 
$$
\mathcal{H}_0 : \beta = 0 \quad \quad \mbox{against} \quad \quad \mathcal{H}_1 :  \beta > 0 \quad \quad \text{with} \quad \lambda(t) = \alpha e^{\beta t}
$$
```{r}
lambda_exp <- function(t, alpha, beta) {
    return(alpha*exp(beta*t))
  }
```

We reject $\mathcal{H}_0$ when $\mathcal{R}_{\alpha} = \left\{ L \ge l \right\}$. 

```{r}
Tstar <- 5

plot(seq(0,Tstar,length.out = 10), lambda_exp(seq(0,Tstar,length.out = 10), alpha = 1, beta = 0), main = "beta = 0 : lambda constant")

plot(seq(0,Tstar,length.out = 10), lambda_exp(seq(0,Tstar,length.out = 10), alpha = 1, beta = 1), main = "beta > 0 : lambda increasing")
```

Let's simulate homogeneous and inhomogeneous Poisson processes in order to validate the test on simulated data. 

```{r}
# In the homogeneous case, beta = 0, so lambda = alpha
Tstar <- 5 
alpha <- 2
lambda <- alpha
PPh1 <- simulPPh(lambda, Tstar)
plot_PP(PPh1)
```
Let's compute the p-value. 
$$
\hat\alpha_n \approx 1 - \Phi(Z^{obs}) \quad \text{where } \Phi \text{ cdf of a } \mathcal{N}(0,1)
$$

```{r}
get_pvalL <- function(PP, Tstar)
{
  n <- length(PP)
  L_obs <- sum(PP/Tstar)
  Z_obs <- (L_obs - n/2)/sqrt(n/12)
  # pvalue : smaller value such that we reject H0
  pval <- 1 - pnorm(Z_obs)
  return(pval)
}
```

```{r}
alpha <- 0.05
pval <- get_pvalL(PPh1, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```
$$
\Lambda(t) = \frac{\alpha}{\beta} \left( e^{\beta t} - 1 \right)
$$

```{r}
lambda_exp <- function(alpha, beta, Tstar) {
  lambda <- function(t) {
    return(alpha*exp(beta*t))
  }
  LAMBDA <- function(t) {
    return((alpha/beta)*(exp(beta*t) - 1))
  }
  
  quantile_function <- function(p) {
    return((1/beta) * log(p*(exp(beta*Tstar) - 1) + 1))
  }
  
  return(c(lambda, LAMBDA, quantile_function))
}
```

```{r}
Tstar <- 3
alpha <- 1 
beta <- 1
expo <- lambda_exp(alpha, beta, Tstar)
lambda <- expo[[1]]
LAMBDA <- expo[[2]]
quantile_function <- expo[[3]]
PPi1 <- simulPPi(LAMBDA, quantile_function, Tstar)
plot_PP(PPi1)
```
 
```{r}
alpha <- 0.05
pval <- get_pvalL(PPi1, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```



# Weibull test 

Test statistic :

$$
Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right )
$$

This test can be used for testing 
$$
\mathcal{H}_0 : \beta=1 \quad \quad \mbox{against} \quad \quad \mathcal{H}_1 : \beta > 1 \quad \quad \text{with} \quad \lambda(t) = \left( \frac{\beta}{\theta}\right) \left(\frac{t}{\theta} \right)^{\beta-1}
$$

```{r}
lambda_weib <- function(t, beta, theta) {
  return((beta/theta)*(t/theta)**(beta-1))
}
```

We reject $\mathcal{H}_0$ when $\mathcal{R}_{\alpha} = \{Z \leq q \}$. 

```{r}
Tstar <- 1

plot(seq(0,Tstar,length.out = 10), lambda_weib(seq(0,Tstar,length.out = 10), beta = 1, theta = 1), main = "beta = 1 : lambda constant")

plot(seq(0,Tstar,length.out = 10), lambda_weib(seq(0,Tstar,length.out = 10), beta = 5, theta = 1), main = "beta > 1 : lambda increasing")
```

```{r}
# In the homogeneous case, beta = 1, so lambda = 1/theta
Tstar <- 1
theta <- 0.1
lambda <- 1/theta
PPh2 <- simulPPh(lambda, Tstar)
plot_PP(PPh2)
```

$\beta Z$ follows a chi-squared with $2n$ degrees of freedom.

$$
\mathcal{R} = \{Z \leq q \} \\
\text{with  } Z = 2 \sum_{i=1}^n \log \left (\frac{T^\star}{T_i} \right ) \\
\text{and } \beta Z \sim \mathcal{X}^2(2n)
$$

P-value computation :

$$
pvalue = \mathbb{P}(Z \leq Z^{obs}) = \Phi_{\mathcal{X^2}}(Z^{obs}) \\
\text{where } \Phi \text{ cdf of a } \mathcal{X}^2(2n)
$$

```{r}
get_pvalZ <- function(PP, Tstar){
  Zobs <- 2*sum(log(Tstar/PP))
  N <- length(PP)
  # pvalue : smaller value such that we reject H0
  pval <- pchisq(Zobs, 2*N)
  return(pval)
}
```

```{r}
alpha <- 0.05
pval <- get_pvalZ(PPh2, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

```{r}
lambda_weibull <- function(beta, theta, Tstar) {
  lambda <- function(t) {
    return((beta/theta)*(t/theta)**(beta-1))
  }
  LAMBDA <- function(t) {
    return((t / theta)**beta)
  }
  quantile_function <- function(p) {
    return(Tstar * p**(1/beta))
  }
  return(c(lambda, LAMBDA, quantile_function))
}
```

```{r}
Tstar <- 1
beta <- 3
theta <- 0.5
weib <- lambda_weibull(beta, theta, Tstar)
lambda <- weib[[1]]
LAMBDA <- weib[[2]]
quantile_function <- weib[[3]]
PPi2 <- simulPPi(LAMBDA, quantile_function, Tstar)
plot_PP(PPi2)
```

```{r}
alpha <- 0.05
pval <- get_pvalZ(PPi2, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```


# F-Test 

Test statistic : 

$$F=\frac{(n-d)T_d}{d(T_n-T_d)}$$

```{r}
lambda <- 2
Tstar <- 20
PPh3 <- simulPPh(lambda, Tstar)
plot_PP(PPh3)
```

```{r}
get_pvalF <- function(PP, Tstar = 1){ # The parameter Tstar here is useless, we add it so that we can use the generic function get_size defined before 
  n = length(PP)
  d = floor(n/2)
  F = (n-d)*PP[d]/(d*(PP[n] - PP[d]))
  pval = 1. - pf(F, 2*d, 2*(n-d))
  return(pval)
}
```

```{r}
alpha <- 0.05
pval <- get_pvalF(PPh3)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

```{r, eval=TRUE}
# lambda increasing by level, return lambda, LAMBDA, quantile_function
lambda_palier <- function(Tstar=10, tau=0.5, a=1, b=2) {
  # a function returning a if t <= tau and b otherwise.
  lambda <- function(t) {
    return(ifelse(t <= tau, a, b))
  }
  # return a function returning the primitive of the precedent one.
  LAMBDA <- function(t) {
    return(ifelse(t <= tau, a*t, a*tau+b*(t-tau)))
  }
  # return a function returning the inverse of the precedent one.
  quantile_function <- function(k) {
    return(ifelse(LAMBDA(Tstar)*k<=tau*a, LAMBDA(Tstar)*k/a, (LAMBDA(Tstar)*k-a*tau)/b + tau))
  }
  return(c(lambda, LAMBDA, quantile_function))
}
```

```{r}
Tstar <- 20
tau <- 2/3
a <- 1
b <- 2
palier <- lambda_palier(Tstar, tau*Tstar, a, b)
lambda <- palier[[1]]
LAMBDA <- palier[[2]]
quantile_function <- palier[[3]]
PPi3 <- simulPPi(LAMBDA, quantile_function, Tstar)
plot_PP(PPi3)
```

```{r}
alpha <- 0.05
pval <- get_pvalF(PPi3, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

With thining method : 

```{r}
PPi4 <- simulPPi_thining(lambda, NULL, Tstar)
plot_PP(PPi4)
```
```{r}
alpha <- 0.05
pval <- get_pvalF(PPi4, Tstar)
cat("P-value : ", pval, "\n")
decision(pval, alpha)
```

# Kolmogrov Smirnov

```{r, eval=TRUE}
library(KScorrect)
get_pvalKS <- function(PP, Tstar = 1) {
  N <- length(PP)
  W <- PP[2:N] - PP[1:N-1]
  D <- LcKS(W, "pexp", nreps = 1999)
  pval <- D$p.value
  return(pval)
}
```

# Size of the tests

```{r}
get_size <- function(K, alpha, get_pval, simulPPh, lamdba, Tstar){
  nb.rejects <- 0
  for(k in 1:K){
    N <- 0
    while(N <= 1){
      PP <- simulPPh(lambda, Tstar)
      N <- length(PP)
    }
    pval <- get_pval(PP, Tstar)
    nb.rejects = nb.rejects + (pval<=alpha)  
  }
  return(nb.rejects/K)
}
```


```{r}
alpha <- 0.05
Tstar <- 5
lambda <- 2
test_size <- get_size(K = 2000, alpha, get_pvalL, simulPPh, lambda, Tstar)
cat("\nSize of the test : ", test_size)
```

```{r}
alpha <- 0.05
Tstar <- 5
theta <- 0.1
lambda <- 1/theta
test_size <- get_size(K = 2000, alpha, get_pvalZ, simulPPh, lambda, Tstar)
cat("\nSize of the test : ", test_size)
```

```{r}
alpha <- 0.05
Tstar <- 20
lambda <- 2
test_size <- get_size(K = 2000, alpha, get_pvalF, simulPPh, lambda, Tstar)
cat("\nSize of the test : ", test_size)
```


# Comparison of the power of the tests 

```{r}
get_power <- function(K, alpha, get_pval, simulPP, LAMBDA, quantile_function, Tstar){
  nb.rejects <- 0
  for(k in 1:K){
    N <- 0
    while(N <= 1){
      PP <- simulPP(LAMBDA, quantile_function, Tstar)
      N <- length(PP)
    }
    pval <- get_pval(PP, Tstar)
    nb.rejects = nb.rejects + (pval<=alpha)  
  }
  return(nb.rejects/K)
}
```

## Some examples of power estimation

```{r}
alpha <- 0.05
Tstar <- 3
test_power <- get_power(K = 2000, alpha, get_pvalL, simulPPi, LAMBDA, quantile_function, Tstar)
cat("\nPower of the test : ", test_power)
``` 

```{r}
alpha <- 0.05
Tstar <- 3
beta <- 4
theta <- 1
weib <- lambda_weibull(beta, theta, Tstar)
lambda <- weib[[1]]
LAMBDA <- weib[[2]]
quantile_function <- weib[[3]]
test_power <- get_power(K = 2000, alpha, get_pvalZ, simulPPi, LAMBDA, quantile_function, Tstar)
cat("\nPower of the test : ", test_power)
``` 

```{r}
alpha <- 0.05
test_size <- get_power(K = 2000, alpha, get_pvalF, simulPPi, LAMBDA, quantile_function, Tstar)
cat("\nPower of the test : ", test_size)
``` 

## Comparative study of test

```{r, eval=TRUE}
# lambda increasing by level, return lambda, LAMBDA, quantile_function
lambda_ln <- function(Tstar=10, beta=1) {
  # a function returning a if t <= tau and b otherwise.
  lambda <- function(t) {
    return(beta*log(1+t))
  }
  # return a function returning the primitive of the precedent one.
  LAMBDA <- function(t) {
    return(beta*((1+t)*log(1+t)-t))
  }
  # quantile function not computed
  return(c(lambda, LAMBDA))
}
```

```{r, eval=TRUE}
# Listes des paramètres des intensités
T_star = c(2, 5, 10)

K = 2000
```

```{r, eval=TRUE}
get_power_all_test <- function(K, alpha, simulPP, LAMBDA, quantile_function, Tstar) {
  nb.rejects.Z <- 0
  nb.rejects.L <- 0
  nb.rejects.F <- 0
  nb.rejects.KS <- 0
  for(k in 1:K){
    # print(k)
    N <- 0
    while(N <= 1){
      PP <- simulPP(LAMBDA, quantile_function, Tstar)
      N <- length(PP)
    }
    pval.Z <- get_pvalZ(PP, Tstar)
    pval.L <- get_pvalL(PP, Tstar)
    pval.F <- get_pvalF(PP, Tstar)
    pval.KS <- get_pvalKS(PP, Tstar)
    nb.rejects.Z = nb.rejects.Z + (pval.Z<=alpha)
    nb.rejects.L = nb.rejects.L + (pval.L<=alpha)  
    nb.rejects.F = nb.rejects.F + (pval.F<=alpha)  
    nb.rejects.KS = nb.rejects.KS + (pval.KS<=alpha)  
  }
  return(c(nb.rejects.Z/K, nb.rejects.L/K, nb.rejects.F/K, nb.rejects.KS/K))
}
```

```{r, eval=TRUE}
# Weibull intensity
beta_weib = c(2, 2, 2)
theta_weib = c(0.6, 0.7, 0.7)

wei_Z  = c()
wei_L  = c()
wei_F  = c()
wei_KS = c()
for(i in 1:3) {
  T = T_star[[i]]
  weib <- lambda_weibull(beta_weib[[i]], theta_weib[[i]], T)
  lambda <- weib[[1]]
  LAMBDA <- weib[[2]]
  quantile_function <- weib[[3]]
  res = get_power_all_test(K, alpha, simulPPi, LAMBDA, quantile_function, T)
  wei_Z = c(wei_Z, res[[1]])
  wei_L = c(wei_L, res[[2]])
  wei_F = c(wei_F, res[[3]])
  wei_KS = c(wei_KS, res[[4]])
}
```

```{r, eval=TRUE}
# Exponential intensity
alpha_exp = c(2, 0.35, 0.009)
beta_exp = c(1, 1, 1)

exp_Z  = c()
exp_L  = c()
exp_F  = c()
exp_KS = c()
for(i in 1:3) {
  T = T_star[[i]]
  exp <- lambda_exp(alpha_exp[[i]], beta_exp[[i]], T)
  lambda <- exp[[1]]
  LAMBDA <- exp[[2]]
  quantile_function <- exp[[3]]
  res = get_power_all_test(K, alpha, simulPPi, LAMBDA, quantile_function, T)
  exp_Z = c(exp_Z, res[[1]])
  exp_L = c(exp_L, res[[2]])
  exp_F = c(exp_F, res[[3]])
  exp_KS = c(exp_KS, res[[4]])
}
```

```{r, eval=TRUE}
# Step function intensity
tau = 0.5
a_pal = c(3, 5, 10)
b_pal = c(6, 10, 20)

pal_Z  = c()
pal_L  = c()
pal_F  = c()
pal_KS = c()
for(i in 1:3) {
  T = T_star[[i]]
  pal <- lambda_palier(T, tau, a_pal[[i]], b_pal[[i]])
  lambda <- pal[[1]]
  LAMBDA <- pal[[2]]
  quantile_function <- pal[[3]]
  res = get_power_all_test(K, alpha, simulPPi, LAMBDA, quantile_function, T)
  pal_Z = c(pal_Z, res[[1]])
  pal_L = c(pal_L, res[[2]])
  pal_F = c(pal_F, res[[3]])
  pal_KS = c(pal_KS, res[[4]])
}
```

```{r, eval=TRUE}
# Logarithmic intensity
beta_ln = c(7.7, 8.7, 12.2)

ln_Z  = c()
ln_L  = c()
ln_F  = c()
ln_KS = c()
for(i in 1:3) {
  T = T_star[[i]]
  ln <- lambda_ln(T, beta_ln[[i]])
  lambda <- ln[[1]]
  LAMBDA <- ln[[2]]
  res = get_power_all_test(K, alpha, simulPPi_thining, lambda, NULL, T)
  ln_Z = c(ln_Z, res[[1]])
  ln_L = c(ln_L, res[[2]])
  ln_F = c(ln_F, res[[3]])
  ln_KS = c(ln_KS, res[[4]])
}
```
# Real data

```{r}
library(evir)
data(danish)
test <- attributes(danish)$times
length(test)
danish <- difftime(test, min(test), units = "days")

# round(test, units = "days")

PPi_danish <- as.numeric(danish)
Tstar <- max(PPi_danish)

plot_PP(PPi_danish)
```


```{r}
# A modifier ! Quel test utiliser ? 
# pval <- get_pval(PPi_danish, Tstar)
# cat("P-value : ", pval, "\n")
# 
# # accept
# if (pval < alpha) {
#   cat("Reject H0 : lambda is inscreasing\n")
# } else {
#   cat("Accept H0 : lambda is constant\n")
# }
```

```{r}
df <- data.frame(matrix(ncol = 12, nrow = 4))
colnames(df) <- c(
  "wei_T1", "wei_T2", "wei_T3",
  "exp_T1", "exp_T2", "exp_T3",
  "pal_T1", "pal_T2", "pal_T3",
  "ln_T1", "ln_T2", "ln_T3")
rownames(df) <- c("Test Weibull", "Test Laplace", "Test F", "Test K.-S.")
df[1,1:12] <- c(wei_Z, exp_Z, pal_Z, ln_Z)
df[2,1:12] <- c(wei_L, exp_L, pal_L, ln_L)
df[3,1:12] <- c(wei_F, exp_F, pal_F, ln_F)
df[4,1:12] <- c(wei_KS, exp_KS, pal_KS, ln_KS)
```

```{r}
df
```