---
title: "Étude du F-test (Snedecor's F-distribution)"
author: "Alexandre Personnic"
date: "November-December 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1 - Theorical construction

First, we define the test statistic we are foing to use. 

Following the observations from Barlow et al (1972, p. 197), we decided to
divide the failure times in two parts, the first $d$ and the last $n-d$, and to
use the statistic $F=(n-d)T_d/(d(T_n-T_d))$ for the test.

Taking $d=\left\lfloor\dfrac{n}{2}\right\rfloor$, if the intensity is
increasing, $T_n - T_d$ should be smaller than for a constant intensity.
Hence an increasing intensity should correspond to a larger value of this ratio.

The random variable $T_d$ (resp. $T_n-T_d$) follows a Gamma distribution with parameters $n$
and $\lambda$ (resp. $n-d$ and $\lambda$). As such, $F$ follows a Snedecor's F-distribution
with $2d$ and $2(n-d)$ degrees of freedom.

\subsection{1.1 - Simulation}

```{r,eval=TRUE}
simulPPh1 <- function(lambda, Tmax) {
  Y <- rpois(1, lambda * Tmax)
  U <- runif(Y, min = 0, max = Tmax)
  return(sort(U))
}
```

```{r,eval=TRUE}
# simulate a homogeneous Poisson process:
lambda <- 2
Tmax <- 10
PPh1 = simulPPh1(lambda, Tmax)


# plot the counting process (with jumps = 1): 
plot(c(0,PPh1),seq(0,length(PPh1)),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times: 
points(PPh1, rep(0,length(PPh1)),type="p")

# link the arrival times with the counts:
lines(PPh1, seq(1,length(PPh1)),type="h",lty=2)
```

```{r, eval=TRUE}
n = length(PPh1)
d = floor(n/2)
F = (n-d)*PPh1[d]/(d*PPh1[n] - PPh1[d])
alpha = 1. - pf(F, 2*d, 2*(n-d))
print(alpha)
```

\section{3 - Inhomogeneous Poisson processes}

Third, we simulate inhomogeneous Poisson processes with given intensity function on a fixed window. 

```{r,eval=FALSE}
simulPPi = function(lambda_fct,Tmax,M) {
  Ti <- simulPPh1(M, Tmax)
  n <- length(Ti)
  U <- runif(n, min = 0, max = M)
  ti <- c()
  for (i in 1:n){
    if (U[i] < lambda_fct(Ti[i])) {
      ti <- c(ti, Ti[i])
    }
  }
  return(sort(ti))
}
```

```{r,eval=FALSE}
Tmax=10
lambda_fct1 <- function(t){
  return(ifelse(t <= 7, 2, 
                ifelse(t > 8, 8, 0)))
  }
M1=10

PPi1 = simulPPi(lambda_fct1, Tmax, M1)

# plot the counting process (with jumps = 1): 
plot(c(0,PPi1),seq(0,length(PPi1)),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times: 
points(PPi1, rep(0,length(PPi1)),type="p")

# link the arrival times with the counts:
lines(PPi1, seq(1,length(PPi1)),type="h",lty=2)
```

```{r, eval=TRUE}
n = length(PPi1)
d = floor(n/2)
F = (n-d)*PPi1[d]/(d*PPi1[n] - PPi1[d])
alpha = 1. - pf(F, 2*d, 2*(n-d))
print(alpha)
```

```{r,eval=FALSE}
Tmax=10
lambda_fct2 <- function(t){
  return(2*t)
  }
M2= 21

PPi2 = simulPPi(lambda_fct2, Tmax, M2)

# plot the counting process (with jumps = 1): 
plot(c(0,PPi2),seq(0,length(PPi2)),type="s",xlab="time t",ylab="number of events by time t")

# add the arrival times: 
points(PPi2, rep(0,length(PPi2)),type="p")

# link the arrival times with the counts:
lines(PPi2, seq(1,length(PPi2)),type="h",lty=2)
```




